---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# valiscale

<!-- badges: start -->
<!-- badges: end -->

The goal of valiscale is to calculate several metrics for quantitative characterization of estimation models performance.

## Installation

You can install the development version of valiscale from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("gabrielblain/valiscale")
```

## Example 1

Got paired daily data (like observed vs. estimated values) and need to aggregate it into common environmental time scales? Here is how to compute averages across daily, 5-day (pentad), and seasonal windows:

```{r aggregate mean}
library(valiscale)
df <- data.frame(
Date = seq(as.Date("2000-01-01"), by = "day", length.out = 9496),
obs = rnorm(9496, mean = 20, sd = 3),
est = rnorm(9496, mean = 20, sd = 3)
)
#Keeping daily:
daily_data <- aggregate_scale(df, scale = "daily", FUN = mean)
head(daily_data)
#pentad (5-day periods):
pentad_means <- aggregate_scale(df, scale = "pentad", FUN = mean)
head(pentad_means)
#seasonal (Austral seasons: DJF, MAM, JJA, SON):
seasonal_means <- aggregate_scale(df, scale = "seasonal", FUN = mean)
head(seasonal_means)
```

Note that averaging at the daily scale produces the same values as the input data.

## Example 2

If you need totals instead of averages, swap `FUN = mean` for `FUN = sum`:
```{r aggregate sum}
df <- data.frame(
Date = seq(as.Date("2000-01-01"), by = "day", length.out = 9496),
obs = rnorm(9496, mean = 20, sd = 3),
est = rnorm(9496, mean = 20, sd = 3)
)
#Keeping daily:
daily_data <- aggregate_scale(df, scale = "daily", FUN = sum)
head(daily_data)
#pentad (5-day periods):
pentad_sum <- aggregate_scale(df, scale = "pentad", FUN = sum)
head(pentad_sum)
#seasonal (Austral seasons: DJF, MAM, JJA, SON):
seasonal_sum <- aggregate_scale(df, scale = "seasonal", FUN = sum)
head(seasonal_sum)
```
Note that summing at the daily scale produces the same values as the input data.

## Example 3

Now let's evaluate model performance across different time scales using the aggregated data from Example 1:

```{r agreement mean}
#Performance for daily values:
agreement_stats(df=daily_data)
#Performance for pentad values:
agreement_stats(df=pentad_means)
#Performance for seasonal values (Austral seasons: DJF, MAM, JJA, SON):
agreement_stats(df=seasonal_means)
```

## Example 4

Overall stats are great, but let's see if model performance varies across different seasons:

```{r agreement mean-seasons}
#Seasonal performance for daily values in the:
agreement_stats(df=daily_data, by_season = TRUE)
#Seasonal performance for pentad values:
agreement_stats(df=pentad_means, by_season = TRUE)
#Seasonal performance for seasonal values (Austral seasons: DJF, MAM, JJA, SON):
agreement_stats(df=seasonal_means, by_season = TRUE)
```

## Example 5

Great. Let's see if model performance varies across different months:

```{r agreement mean-month}
# Monthly performance for daily values
agreement_stats(df = daily_data, by_month = TRUE)
# Monthly performance for pentad values (5-day averages)
agreement_stats(df = pentad_means, by_month = TRUE)
# Monthly performance for seasonal values
# Note: Since seasonal values are already aggregated into 3-month blocks,
# attempting to compute monthly stats here will (correctly) throw an error/warning.
agreement_stats(df = seasonal_means, by_month = TRUE)
```
